---
title: "DATA 607 - Assignment5 [Tidy Data]"
author: "Simon63"
date: "March 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####Package Installs
install.packages("tidyr")  
install.packages("dplyr")
install.packages("ggplot2")

```{r lib, warning=FALSE, message=FALSE}
require(tidyr)
require(dplyr)
require(ggplot2)
require(tibble)
```

###Load the source .CSV file
```{r load_data}
df <- read.csv("https://raw.githubusercontent.com/simon63/Assignment5/master/flights-wk5.csv",
               header = T, stringsAsFactors = F, fileEncoding = "UTF-8-BOM", check.names = F)
# header = T - using this option to use the first line for column names
# check.names = F - using this option to avoid "Los Angeles" being converted to "Los.Angeles"
# fileEncoding = "UTF-8-BOM" - using this option to avoid loading the "\ufeff" encoding character
df
```


###Tidy Data - Assign missing column names
```{r tidy_missing_cols}
# Missing column names must be assigned before tbl_df() can be used
names(df)[1:2] <- c("Airline", "n_type")

# When attempting to use tbl_df(), I noticed a pop-up description, saying:
#   "Deprecated: please use tibble::as_tibble() instead."
df <- as_tibble(df)
df
```

###Tidy Data - Convert columns with city names into a single variable, "city"
```{r tidy_gather}
# Using gather() to do the conversion and 
# also filtering out rows with NA value in the new "n" column
t <- gather(df, "city", "n", 3:7) %>% filter(!is.na(n))
t
```

###Tidy Data - Assign missing values for Airline in even rows
```{r tidy_missing_airline, warning=FALSE, message=FALSE}
# Assign the Airline values from the "Odd" rows into the "Even" rows
t[(c(1:length(t)) %% 2) == 0,1] <- t[(c(1:length(t)) %% 2) != 0,1]
t
```

###Tidy Data - Use spread() function to separate "on time" and "delayed" observations into new columns
```{r tidy_spread}
# Spread would not work without the prior step of assigning missing Airline values
t <- spread(t, n_type, n)
t
```


###Tidy Data - Mutate by adding two new columns, "% on time" and "% delayed"
```{r tidy_mutate}
t <- mutate(t, pct_delayed = round((delayed / (delayed + t$'on time')) * 100, 2), pct_on_time = 100 - pct_delayed)
t
```

###Plot Data - Comparing delays for the two airlines
```{r plot, echo=FALSE}
g <- ggplot(data = t, aes(x = city, y = pct_delayed, fill = Airline)) + geom_col(position = "dodge") + labs(x = "City", y ="% Delayed")
g
```
